Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-06-19 cordillera-cycle.tex
Add corrections from Irina (part 3).

On the Interior Plateau of south-central British Columbia, both simulations
produce a retreat of the ice margin towards the north-east
(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
and stratigraphical record for ancient pro-glacial lakes dammed by the
retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
simulations differ in the mode of retreat. The GRIP simulation yields an active
retreat with basal sliding towards the ice margin to the south, whereas the
EPICA simulation produce negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
set of glacial lineations which indicate a substantial eastward flow component
of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system observed in the
geomorphology predates the deglaciation, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by moisture intruding inland from the
west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap is perhaps associated
with the distinct glacial erosional imprint on the landscape of the Skeena
Mountains.

@@

% Author contributions
\section*{Author contributions}
J.~Seguinot ran the simulations and wrote most of the manuscript; I.~Rogozhina
guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
the interpretation and comparison of model results against geological evidence.
All co-authors contributed to improve the text.

% Acknowledgements
\begin{acknowledgements}
Foremost, we would like to thank Shawn Marshall for providing a detailed,
constructive analysis of this study during J.~Seguinot's Ph.D. defence
(September 2014). His comments were used to improve the model set-up.
We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
providing constant help and development on PISM. This work was supported by the
Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
by the Swedish National Infrastructure for Computing (SNIC) allocation
no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
Center (NSC).
\end{acknowledgements}

>>>>

On the Interior Plateau of south-central British Columbia, both simulations
produce a retreat of the ice margin towards the north-east
(Fig.~\ref{fig:deglac}), a pattern which is validated by the geomorphological
and stratigraphical record for ancient pro-glacial lakes dammed by the
retreating ice sheet \citep{Perkins.Brennand.2014}. However, the two
simulations differ in the mode of retreat. The GRIP simulation yields an active
retreat with basal sliding towards the ice margin to the south, whereas the
EPICA simulation produces negligible basal sliding on the plateau during deglaciation
(Fig.~\ref{fig:lastflow}). Yet, the Interior Plateau also hosts an impressive
set of glacial lineations which indicate a substantial eastwards flow component
of the Cordilleran ice sheet \citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, a second explanation for the incongruent
results could be that the Interior Plateau lineation system observed in the
geomorphology predates the deglaciation, as perhaps indicated by some eskers
that appear incompatible with these glacial lineations
\citep[Fig.~9]{Margold.etal.2013a}. Finally, a third explanation could be that
local geothermal heat associated with volcanic activity on the Interior Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and Mackenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by the moisture intruding inland from the
west through a topographic breach in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the pre-LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap perhaps explains
the distinct glacial erosional imprint observed on the landscape of the Skeena
Mountains.

@@

% Author contributions
\section*{Author contributions}
J.~Seguinot ran the simulations; I.~Rogozhina
guided experiment design; A.~P.~Stroeven, M.~Margold and J.~Kleman took part in
the interpretation and comparison of model results against geological evidence.
All authors contributed to the text.

% Acknowledgements
\begin{acknowledgements}
Foremost, we would like to thank Shawn Marshall for providing a detailed,
constructive analysis of this study during J.~Seguinot's Ph.D. defence
(September 2014). His comments were used to improve the model set-up.
We are very thankful to Constantine Khroulev, Ed Bueler and Andy Aschwanden for
providing constant help and development with PISM. This work was supported by the
Swedish Research Council~(VR) grant no.~2008-3449 to A.~P.~Stroeven and by the
German Academic Exchange Service~(DAAD) grant no.~50015537 and a Knut and Alice
Wallenberg Foundation grant to J.~Seguinot. Computer resources were provided
by the Swedish National Infrastructure for Computing (SNIC) allocation
no.~2013/1-159 and~2014/1-159 to A.~P.~Stroeven at the National Supercomputing
Center (NSC).
\end{acknowledgements}
